Microlensing as a possible probe of the internal structure of quasars

### 2015-07-12 paper1/code.tex
editing

We now go beyond the case of an ideal fold, and consider microlensing
by random distributions of point lenses.  This requires numerical
computation of lightcurves, for which we use the microlensing code by
\cite{1999A&A...346L...5W}. The code is based on the hierarchical tree
technique to calculate the gravitational effect of the lensing
masses. The underlying idea is that when the thin lens approximation
is valid, the a microlensing distribution can be reduced to three
planes, the observer plane, the lens plane and the source plane. The
actual effect, namely the distortion of light rays emitted from a
distant source, on the path to the observer, is numerically computed
by through back-tracing: lightrays are shot backward from the observer
to lens plane, where they become deflected according to the angle
dependent on the 2D mass distribution in the lens plane.
\begin{equation}
\vec{\alpha}_{i}=\frac{4G}{c^{2}} \frac{D_{LS}}{D_{S}}\sum_{j=1}^{n}M_j \frac{\vec{r}_{ij}}{r^2_{ij}} 
\end{equation}  
The choice for reversing the photon motion is motivated by computational resources considerations. 
For the computation of the individual deflection angles the tree concept is used. The positions of all lensing masses are sorted into a 2-dimensional grid (in the lens plane). Which is subdivided into 4 smaller squares recursively until every cell contains only one mass. For the actual calculations of the deflection angle not every mass gravitational influence is treated equally. Instead it is made use of the fact that the influence of lenses on the considered light ray falls off with $r^{-1}$ from the point where the ray hits the lens plane. Hence masses situated at higher distances, can be grouped together in the calculation of their gravitational potential, using multipole moments. With this approach the amount of computation time is reduced. \\

@@

When generating the caustic map used in our analysis presented in the second part of the article only two equal point like masses were used as lenses, to produce the magnification map depicted in figure \ref{fig:magnification_map}. The magnification maps calculated by the Microlensing code are in units of Einstein radii, hence they are scale free. Physical scale is introduced to them by multiplying the pixels by the correct distance factor. The Einstein angle is given by   
\begin{equation}
\theta_{E}=\sqrt{\frac{4GM}{c^{2}} \frac{D_{LS}}{D_{S}D_{L}}} 
\end{equation}  
and therefore depends on the four parameters $M,D_{S},D_{L},D_{LS}$. For the cosmological application or quasar lensing $D_{L}$ is of order 100 Mps to Gpc and $D_{S}$ of order Gpc.\\   
The scale of the magnification map can be converted into a physical length by multiplying it by a factor of $ \sqrt{\frac{D_{LS}}{D_{S}D_{L}}}$.\\\\
For the next step, the computation of the actual lightcurve, the code was modified to also allow for crescent shaped images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$ denotes the outer radius of the crescent and $R_n$ the inner one. The orientation of the source image with respect to the magnification map is specified by the parameters $a$ and $b$ as the shift of the center of the inner disk from the center of the outer disk in $x$- and $y$-direction respectively. In the original version of the code, gaussian and disk shaped images where already implemented. Those are completely characterized by the single parameter $R_p$. The values of the parameters are specified in pixel units corresponding to the magnification map. Further one needs to specify the start and end point coordinates of the path, which the center of the source image follows through the magnification map (see the depiction in figure \ref{fig:magnification_map}). Hence the points along this path are specified through the number of timesteps for which the computation of the brightness is to be carried out. Those points correspond to the actual measurement of the brightness of an object in the observational case. For each timestep the two-dimensional convolution of equation \ref{eqn:ft2d} is carried out numerically for the position of the source on the magnification map. \\
For the purpose of this analysis it was desirable to mimic the analytical behaviour of a simple fold as much as possible, for comparing the numerical result with the analytical one, therefore the path of the source was chosen, so that it intersects the border of the caustic perpendicularly, and on a point where the border is a fold caustic.

>>>>

We now go beyond the case of an ideal fold, and consider microlensing
by random distributions of point lenses.  This requires numerical
computation of lightcurves, for which we use the microlensing code by
\cite{1999A&A...346L...5W}.

The code is based on the hierarchical tree technique to calculate the
gravitational effect of the lensing masses. The underlying idea is
that when the thin lens approximation is valid, the a microlensing
distribution can be reduced to three planes, the observer plane, the
lens plane and the source plane. The actual effect, namely the
distortion of light rays emitted from a distant source, on the path to
the observer, is numerically computed by through back-tracing:
lightrays are shot backward from the observer to lens plane, where
they become deflected according to the angle dependent on the 2D mass
distribution in the lens plane.  The choice for reversing the photon
motion is motivated by computational resources considerations.  For
the computation of the individual deflection angles the tree concept
is used. The positions of all lensing masses are sorted into a
2-dimensional grid (in the lens plane). Which is subdivided into 4
smaller squares recursively until every cell contains only one
mass. For the actual calculations of the deflection angle not every
mass gravitational influence is treated equally. Instead it is made
use of the fact that the influence of lenses on the considered light
ray falls off with $r^{-1}$ from the point where the ray hits the lens
plane. Hence masses situated at higher distances, can be grouped
together in the calculation of their gravitational potential, using
multipole moments. With this approach the amount of computation time
is reduced. \\

@@

When generating the caustic map used in our analysis presented in the
second part of the article only two equal point like masses were used
as lenses, to produce the magnification map depicted in figure
\ref{fig:magnification_map}. The magnification maps calculated by the
Microlensing code are in units of Einstein radii, hence they are scale
free. Physical scale is introduced to them by multiplying the pixels
by the correct distance factor. The Einstein angle is given by
\begin{equation}
\theta_{E}=\sqrt{\frac{4GM}{c^{2}} \frac{D_{LS}}{D_{S}D_{L}}} 
\end{equation}  
and therefore depends on the four parameters
$M,D_{S},D_{L},D_{LS}$. For the cosmological application or quasar
lensing $D_{L}$ is of order 100 Mps to Gpc and $D_{S}$ of order
Gpc.\\ The scale of the magnification map can be converted into a
physical length by multiplying it by a factor of
$\sqrt{\frac{D_{LS}}{D_{S}D_{L}}}$.\\\\ For the next step, the
computation of the actual lightcurve, the code was modified to also
allow for crescent shaped images specified through the parameter set
$R_p,R_n,a,b$. Here $R_p$ denotes the outer radius of the crescent and
$R_n$ the inner one. The orientation of the source image with respect
to the magnification map is specified by the parameters $a$ and $b$ as
the shift of the center of the inner disk from the center of the outer
disk in $x$- and $y$-direction respectively. In the original version
of the code, gaussian and disk shaped images where already
implemented. Those are completely characterized by the single
parameter $R_p$. The values of the parameters are specified in pixel
units corresponding to the magnification map. Further one needs to
specify the start and end point coordinates of the path, which the
center of the source image follows through the magnification map (see
the depiction in figure \ref{fig:magnification_map}). Hence the points
along this path are specified through the number of timesteps for
which the computation of the brightness is to be carried out. Those
points correspond to the actual measurement of the brightness of an
object in the observational case. For each timestep the
two-dimensional convolution of equation \ref{eqn:ft2d} is carried out
numerically for the position of the source on the magnification
map. \\ For the purpose of this analysis it was desirable to mimic the
analytical behaviour of a simple fold as much as possible, for
comparing the numerical result with the analytical one, therefore the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.

### 2015-07-12 paper1/fitting.tex
editing

\begin{figure}
\centering
\includegraphics[width=0.9\hsize]{plots/data_cc.eps}
\includegraphics[width=0.9\hsize]{plots/data_dd.eps}
\includegraphics[width=0.9\hsize]{plots/data_gg.eps}
\caption{\label{fig:mockdata} Three different mock datasets with
  three different parameterisations fitted to each --- crescent (red),
  uniform-disk (blue) and Gaussian disk (green).}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{plots/Rp4all.eps}
\caption{\label{fig:mcmc} Posterior probability distribution for the
  source size $R_p$.  The vertical line is the correct value. Same
  color represeents the same dataset whereas same line style
  corresponds to same model fit.  The legend gives the reduced
  $\chi^2$ of the best fit in each case.}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{plots/Rhalf_RnRp.eps}
  \includegraphics[width=0.9\hsize]{plots/aRp_bRp.eps}
\caption{\label{fig:crescentfit} The 2$\sigma$ contours (or error ellipses)
  of the crescent model when fitting with the three different
  datasets.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\hsize]{plots/burnin_cg.eps}
\caption{\label{fig:burnin} The $\chi^2$ of each step of the full MCMC
  for one case.}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=0.9\hsize%,bb=0 0 576 432
                ]{plots/data_cc.eps}
\includegraphics[width=0.9\hsize%,bb=0 0 576 432
                ]{plots/data_dd.eps}
\includegraphics[width=0.9\hsize%,bb=0 0 576 432
                ]{plots/data_gg.eps}
\caption{\label{fig:mockdata} Three different mock datasets with
  three different parameterisations fitted to each --- crescent (red),
  uniform-disk (blue) and Gaussian disk (green).}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize%,bb=0 0 576 432
                  ]{plots/Rp4all.eps}
\caption{\label{fig:mcmc} Posterior probability distribution for the
  source size $R_p$.  The vertical line is the correct value. Same
  color represeents the same dataset whereas same line style
  corresponds to same model fit.  The legend gives the reduced
  $\chi^2$ of the best fit in each case.}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize%,bb=0 0 576 432
                  ]{plots/Rhalf_RnRp.eps}
  \includegraphics[width=0.9\hsize%,bb=0 0 576 432
                  ]{plots/aRp_bRp.eps}
\caption{\label{fig:crescentfit} The 2$\sigma$ contours (or error ellipses)
  of the crescent model when fitting with the three different
  datasets.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\hsize%,bb=0 0 576 432
                ]{plots/burnin_cg.eps}
\caption{\label{fig:burnin} The $\chi^2$ of each step of the full MCMC
  for one case.}
\end{figure}

### 2015-07-12 paper1/theory.tex
editing

\subsection{Basic microlensing theory}

We start by introducing in a succinct manner the gravitational lensing theory that is relevant for microlensing in general and for the scope of the
present paper in particular. For a detailed presentation of the theory the reader is referred to the books and articles that represent the references
of the current section (i.e. \cite{Schneider_book}). The general microlensing equation can be written as:
\begin{equation}
\vec{y} =   
 \begin{pmatrix}
  1-\gamma - k & 0 \\
  0 & 1 + \gamma -k \\
 \end{pmatrix} 
\vec{x} - \sum_{i=1}^{n} m_i \frac{\vec{x} - \vec{x_i}}{\left( \vec{x} - \vec{x_i} \right)^2}.
\label{eqn:microlensingequation}
\end{equation}

The equation is valid for a set of $\it{n}$ point (Schwarzschild) lenses placed at coordinates $\vec{x_i}$. $\vec{y}$ represents the coordinates in the source plane for the respective ray, while $\vec{x}$ represents the coordinates in the lens plane. Furthermore, $\gamma$ and $k$ are the shear and surface mass density. For the lens mapping, a Jacobian can be defined:
\begin{equation} 
A(\vec{x}) = 
 \begin{pmatrix}
  \frac{\partial{y_1}}{\partial{x_1}} & \frac{\partial{y_1}}{\partial{x_2}} \\
  \frac{\partial{y_2}}{\partial{x_1}} & \frac{\partial{y_2}}{\partial{x_2}} \\
 \end{pmatrix}. 
\end{equation}
The determinant of $A(\vec{x})$ is related to the local magnification factor through the equation:

\begin{equation}
\mu(\vec{x}) = \left| \frac{1}{det(A(\vec{x}))} \right|
\label{eqn:mu_x}
\end{equation}

According to this equation, from an infinitesimally small source positioned at $\vec{x}$ in the lens plane an observed will receive a flux $\mu(\vec{x}) dF$ instead of the flux $dF$ that would be
observed in the absence of the gravitational lens. In other words the source will dim or brighten depending on whether $\mu(\vec{x})$ is sub-unitary or supra-unitary.\\

In general, for a single infinitesimally small source more than one image can be observed, each with magnification $\mu_j$. For microlensing events the angular sizes of the
images cannot be resolved. The only observable is the flux to which all nearby images contribute. A total magnification
\begin{equation}
 \mu_t = \sum_{j} \mu_j
\end{equation}
 can be defined and account for an increase/decrease of brightness from all images.

\subsection{Magnification of a point source near a fold}

The determinant of the Jacobian $A$ can have both positive and negative values depending in which region it is computed. Such regions are well defined and separated
by critical curves where the determinant vanishes. The lens equation maps the critical curves of the lens plane into caustics in the source plane. An example of the caustic 
shape can be seen in figure \ref{fig:magnification_map}.\\

@@

Along these critical curves in the lens plane or caustics in the source plane,  the magnification $\mu$ is infinite, a result that simply follows from equation \ref{eqn:mu_x}. This statement holds only for infinitesimally small sources. For extended sources the maximum amplification is finite since only an infinitesimally small area of the source
overlaps with the caustics \citep{1986BAAS...18T.907S}.

The matrix $A$ at the coordinates of the caustic can have rank 1 or 0. If the rank is 1 then the coordinates belong to {\em fold\/} or {\em cusp\/} catastrophes (\cite{1992A&A...260....1S}, \cite{2008mmc..confE...8D}). For the purpose 
of this paper we are interested only in the behaviour around {\em fold\/} catastrophe. A condition necessary to distinguish between a fold and cusp singularity is that the
eigenvector of $A$ corresponding to the null eigenvalue is not a tangent vector of the critical curve (\cite{1992A&A...260....1S}, \cite{2008mmc..confE...8D}).

\begin{figure}
\includegraphics[width = .49\textwidth]{plots/infinite_fold.eps}
\caption{\label{fig:infinite_fold}  Source profiles for gaussian, disk, crescents (right) and magnification map for an infinite fold (left). Objects have the same $S_{0}$ and $r_{1/2}$. }
\end{figure}

The behaviour of the images of a point source and their corresponding magnifications near such a caustic has been thoroughly studied in the past. The reader who is interested in 
understanding the details related to this subject is referred to papers such as \cite{1986ApJ...310..568B}, \cite{1992A&A...260....1S}, \cite{2002ApJ...574..970G}, \cite{2002ApJ...580..468G}. For the present paper we only 
summarize the common result:
\begin {equation}
 \mu_{t}(p) = \mu_0 + C_0 \frac{1}{\sqrt{p}} \Theta(p).
\end {equation}
As the previous equation describes the magnification of a point source near a caustic is equal to the sum of the magnification due to other reasons $\mu_0$, assumed to be locally constant,  and a decrease with
the square root of the distance from the fold. The later term becomes activated only after the source enters the region interior to the caustic curves when the values of the stepfunction $\Theta(p)$ become unity. The proportionality
constant $C_0$  is sensible only to the local behaviour of Fermat's potential and units used for the lens plane coordinates $\vec{y}(p,q)$.

\subsection{Magnification of an extended source near a fold}

@@

For a microlensing event the lightcurve can be written for an undefined source shape as:

\begin{equation}
 F(t) = \int_{-\infty}^\infty \int_{-\infty}^\infty S_{2D}(p-p_s(t), q-q_s(t)) \mu_t(p) \mathrm{d}q \mathrm{d}p
 \label{eqn:ft2d}
\end{equation}

In order to build the previous equation we have considered that the time dependency of the flux $F$ is given only by the motion of the source with respect to a fixed caustic. Therefore the only time dependent quantities in the right hand side of the equation
are the coordinates of the center of the source $p_s,q_s$ and by construct $S_{2D}$.

@@

\subsection{Crescent source with constant surface brightness distribution}

>>>>

We start by introducing in a succinct manner the gravitational lensing
theory that is relevant for microlensing in general and for the scope
of the present paper in particular.  More detailed presentation of the
theory can be found in several references, such as
\cite{2001stgl.book.....P}.

\subsection{Microlensing essentials}

The gravitational lens equation
\begin{equation}
\vec\beta = \vec\theta - \vec\alpha(\vec\theta)
\label{eqn:lens}
\end{equation}
relates the apparent sky position $\vec\theta$ of a light source to
its true but unobservable sky position $\vec\beta$ through the bending
angle $\vec\alpha$.  The latter is an integral 
\begin{equation}
\vec\alpha = (1+z_{_L})\frac{D_S-D_L}{D_SD_L} \frac{4G}{c^2}
\int \Sigma(\vec\theta')
\frac{\vec\theta-\vec\theta'}{|\vec\theta-\vec\theta'|^2}\,
d^2\vec\theta'
\end{equation}
depending on the projected density $\Sigma(\vec\theta)$ (kg/steradian)
of the lens, the lens redshift $z_{_L}$ and the comoving distances
$D_L$ and $D_S$ to the lens and source.  The derivative of the
apparent position with respect to the source position
\begin{equation}
M(\vec\theta) =
\left(\frac{\partial\beta}{\partial\theta}\right)^{-1}
\label{eqn:magnif-matrix}
\end{equation}
is known as the magnification matrix, and its determinant
\begin{equation}
\mu(\vec\theta) = \det|M(\vec\theta)|
\end{equation}
is the brightness amplification of an image of a point source.  In
other words, the source will brighten or dim according to whether
$\mu(\vec\theta)$ is more or less than unity.  If there are multiple
mages at distinct but not observationally resolved $\vec\theta_i$ from
the same $\vec\beta$, a total brightness amplification
of\begin{equation} \mu_{\rm total} = \sum_{i} \mu(\vec\theta_i)
\end{equation}
applies. It is possible for the magnification to become formally
infinite, as a result of an eigenvalues of the magnification matrix
(\ref{eqn:magnif-matrix}) becoming infinite.  This generically happens
on curves on the $\vec\theta$ plane, known as {\em critical curves}.
Mapping a critical curve to the source plane, through the lens
equation (\ref{eqn:lens}), to the source plane $\vec\beta$ gives the
so-called caustics.  Caustics can appear in optical system, not just
gravitational lensing.  For a point source, caustics are singularities
of the magnification; for finite-size sources caustics correspond to
high and sharply changing magnification.

Caustics are important in all forms of lensing with multiple images,
but they have a special significance for lens quasars, first pointed
out by \cite{1979Natur.282..561C}.  The granularity of the mass
distribution $\Sigma(\vec\theta)$ due to individual stars produces a
caustic network on the scale of $\sqrt{GM_\odot D_L/c^2}$ or
micro-arcsec.  Extended sources wash out this micro-caustic structure,
but the optical-continuum source of quasars is even smaller.  Hence
sharp changes in the brightness of lensed quasars are observable, as
they cross the micro-caustics \citep[e.g.,][]{2012A&A...544A..62S}.

@@

\subsection{Magnification near a fold caustic}

A simple example of a caustic network is shown in
Figure~\ref{fig:magnification_map}.  There are two general categories
of caustics in gravitational lensing, cusps and fold, and examples of
both kinds can be seen in this figure.  Magnification near a caustic
has universal properties, independent of the system and has been
extensively studied
\citep{1986ApJ...310..568B,1992A&A...260....1S,2002ApJ...574..970G,2002ApJ...580..468G}.
In particular, at distance $p$ from a fold caustic
\begin {equation}
 \mu(p) = \mu_0 + C_0 \frac{1}{\sqrt{p}} \Theta(p).
\end {equation}
Here the magnification of a point source near a caustic is equal to
the sum of the magnification due to other reasons $\mu_0$, assumed to
be locally constant, and a decrease with the square root of the
distance from the fold. The later term becomes activated only after
the source enters the region interior to the caustic curves when the
values of the stepfunction $\Theta(p)$ become unity. The
proportionality constant $C_0$ depends on the local conditions in the
vicinity of the caustic.

@@

\begin{figure}
\includegraphics[width = .49\textwidth]{plots/infinite_fold.eps}
\caption{\label{fig:infinite_fold} Source profiles for Gaussian, disk,
  crescents (right) and magnification map for an infinite fold
  (left). Objects have the same $S_{0}$ and $r_{1/2}$. }
\end{figure}

For a microlensing event the lightcurve can be written for an
undefined source shape as:
\begin{equation}
 F(t) = \int_{-\infty}^\infty \int_{-\infty}^\infty S_{2D}(p-p_s(t), q-q_s(t)) \mu_t(p) \mathrm{d}q \mathrm{d}p
 \label{eqn:ft2d}
\end{equation}
In order to build the previous equation we have considered that the
time dependency of the flux $F$ is given only by the motion of the
source with respect to a fixed caustic. Therefore the only time
dependent quantities in the right hand side of the equation are the
coordinates of the center of the source $p_s,q_s$ and by construct
$S_{2D}$.

@@

\subsection{Crescent source with constant surface brightness distribution}\label{subsec:crescent}
